Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-06-02 paper/manuscript.tex

Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.

Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
\begin{equation} \label{eq:DGfromKAB}
    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
\end{equation}
Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.

\paragraph{Handling the standard state in absolute free energy calculations.}
For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.

For binding, however, the situation is more complex and requires special care.
Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
\begin{equation}\label{eq:restraint-correction}
    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
\end{equation}
where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.

\paragraph{Several choices of restraints are possible.}
In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.

Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.

In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
Additionally, such restraints would typically keep the ligand from exploring alternative binding modes. This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.

Many choices of restraints involve selecting reference atoms.
Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example. Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.

More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms. Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand. Fu et al. suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}. Salari et al. proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}. This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction. This necessitates an additional step where the DBC restraints are released to a single harmonic restraint. Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.

>>>>

An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species. Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested. Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units. A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).

In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex, Eq.~(\ref{eq:intra-zRL-def}).  The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average. To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom. Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand. To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
\begin{equation} \label{eq:Vsite-def}
    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
\end{equation}
%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
%for dynamical variables rather than integrated quantities
which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent. The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart. The result is
\begin{equation}\label{eq:DG0-binding-constant-def-average}
K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
\end{equation}
where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
\begin{equation}\label{eq:pert-energy-def}
\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
\end{equation}
is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$. The ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble. Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
\end{equation}
which, expressed in free energy units, yields
\begin{eqnarray}\label{eq:DG0-def-Vsite}
  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
\end{eqnarray}

The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$. The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration. The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$. As seen above, the ideal term is a fundamental aspect originating from the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.

Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system. The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.

\paragraph{Handling the standard state in free energy calculations.}
As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward. Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations. These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions. Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.

As discussed above, however, the situation for binding is more complex and requires special care. The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor. Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy. Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates. Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, in molecular dynamics applications, the indicator function is often approximated by a flat-bottom restraint~\cite{chen2007can} (see below). Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.

Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions. The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs). However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$. Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}. Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions. Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined. The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.

\paragraph{Several choices of binding site restraints are possible.}
A variety of restraint types are commonly used in alchemical binding free energy calculations. Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations. Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process. In particular, they must be present unchanged at the unbound and bound states. In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies. Hence, they are typically turned on during an initial stage, starting from the uncoupled state, and are turned off in the final stage, near the bound state. In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing standard state terms. However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are actually auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.

Binding site restraints implement the $I(\zeta)$ indicator function. They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor. Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate. The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice. The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand. In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions. However, practical considerations may be important. For example, some choices of Boresch-style anchor atoms can be numerically unstable. Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail. Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.

Recently, more restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms. Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand. Fu et al. suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}. Salari et al. proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}. DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms. However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints. These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.

As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}. Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).

Auxiliary restraints take a variety of forms, such as simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use} as well as flat-bottom restraints. The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints. For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose. Some auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values. However, more computational effort is required to compute the free energy when the restraints are turned on and off. Additionally, such restraints can limit the exploration of alternative binding modes. This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
